\documentclass[10pt,paper=letter]{scrartcl}
\usepackage[thin,alttitle]{cjquines}

\begin{document}

\subsubsection*{Before}

\begin{itemize}
  \item Way before: replace audio cord in math lab, check any remaining homework.
  \item Before: prepare session 12 handouts, slides, audio, seating arrangements, exam permits, final exam, answer sheets, final exam solutions, certificates, course evaluations.
\end{itemize}

\subsubsection*{Morning lecture}

\begin{itemize}
  \item Before anyone arrives, setup files: slides, audio. Install Sumatra on computer. Make air conditioning as cold as possible as preparation for the simulation.
  \item As people arrive, collect homework for week $6$: problem sets and mock test. Hand out \verb+prime12-exercise.pdf+ and give instructions for people to answer. Hand out exam permits.
  \item Run through \verb+prime12-slides-pre.pdf+. Last year's \verb+primex.pdf+ as reference. Stop at last slide of Elimination, slide 16: ``What non-zero value of $x$\dots''.
  \item Check \verb+prime12-exercise.pdf+. Answer key: BBABB ABDBD CCDBD CDCCC CDBBC BDCDA.
  \item Continue run through \verb+prime12-slides-pre.pdf+. Sine wave speech is on last slide ``Can we be\dots'', on bullet point ``Exposure produces recognition.''.
  \item ``Finally, let's do some problems to warm up before the finals.'' Run through \verb+prime12-slides.pdf+.
\end{itemize}

\subsubsection*{Qualifying simulation}

\begin{itemize}
  \item Check seats and exam permits. Introduce the exam, saying it will be for two and a half hours. Say that they have five minutes to go to the washroom.
  \item Ask to place things in front. Only things allowed to bring are food, ruler, pencil, compass, pen, sharpener, eraser. Say no calculators, graphing paper, protractors, bathroom breaks. Require use of pencil.
  \item Distribute answer sheets and instruct how to fill out details. Distribute scratch paper and say that more cannot be requested. Then distribute questionnaires face down, and ask not to open until exam starts.
  \item Write how much time left in thirty-minute intervals while checking homework and eating lunch. Announce when exactly fifteen minutes are left.
  \item Announce time is up. Collect answer sheets. Distribute solutions, announce they have thirty minutes for disputes, lunch.
\end{itemize}

\subsubsection*{Lunch}

\begin{itemize}
  \item Lunch for an hour. Check papers and input grades in sheet, wait for disputes and correct if any.
  \item Prepare course evaluations, certificates.
\end{itemize}

\subsubsection*{Grades and closing}

\begin{itemize}
  \item Distribute course evaluations and give instructions. Call people one-by-one and show them their grading spreadsheet and letter grades.
  \item Collect course evaluations, give certificates in final ceremony. Then give final reminders: eat well, lots of sleep, don't get sick.
  \item Logistics: what to wear, time and place to meetup, bring pen and pencil, sharpeners, erasers, rulers, compasses, pocket money. 
\end{itemize}

\end{document}
